---
title: "Data analysis"
output: html_document
date: "2023-07-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data

```{r data, warning=FALSE, message=FALSE}
setwd("E:/Cours/SupAgro/Stage M1/Analyses/Romann/Data")

#packages
library(data.table)
library(dplyr)
library(ggplot2)
library(factoextra)
library(car)

rm(list=ls()) #to clear the memory if needed

##### DATA #####

#importation of reads table
seqtab_final<-readRDS("seqtab_final.rds")
tab_final<-t(seqtab_final)
tab_final2<-tab_final[,-4] #elimination of sample 104 because it looks like an error
colnames(tab_final2)<-paste0("EPO_", colnames(tab_final2))

#creation of binary (presence/absence) table
bin_final<-tab_final2
bin_final[bin_final>0]<-1

#importation of the taxa table
tax_final<-readRDS("tax_final.rds")


```

```{r sums}

merged<-merge(tax_final, tab_final2, by=0) #merged table with taxa and nb of reads

#binarytable
presence<-merge(tax_final, bin_final, by=0) #merged table with taxa and presence/absence

#sum of reads for an asv
readsum<-rowSums(merged[,9:187])
totreads<-sum(readsum) #8737989

#nb of samples with detection for an asv
presencesum<-rowSums(presence[,9:187])


#table with interesting values
values<-merged[,1:8]
values<-cbind(values,readsum,presencesum)

#nb of reads per sample (histogram)
sumsample<-colSums(merged[,9:187])
hist(sumsample, col="steelblue")
abline(v = mean(sumsample), col = 'red') #adding the mean vertical line

#nb of species per sample (histogram)
presencesamplesum<-colSums(presence[,9:187])
hist(presencesamplesum, col="steelblue")
abline(v = mean(presencesamplesum), col = 'red') #adding the mean vertical line


```

```{r group, message=FALSE}
#to have a look at the data ! We can change the "group by" parameter
groupvalues <- values %>% group_by(Phylum, Genus) %>%
  summarise(total_reads = sum(readsum), maxnbpresence = max(presencesum), nbvar = n())

```

The total number of reads among all asv is `8737989`. Genus Blumeria (powdery mildew agent) represents around `46%` of the reads.

```{r aggregation}

####### Aggregation of reads table
setDT(merged)

group_cols <- c("Phylum,Genus")  #We group by Genus. 
sample_cols <- names(merged)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- merged[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified Genus
aggreads <- result[complete.cases(result[, Genus]), ]

newnames<- aggreads$Genus
aggreads2 <- data.frame(aggreads[, -1:-2], row.names = newnames)
colnames(aggreads2)<-colnames(aggreads[,3:181]) #to get rid of the newly added x before sample name

#Creation of the aggregated table of presence
aggbin<-aggreads2
aggbin[aggbin>0]<-1





#nb of reads per genus
sumgen<-rowSums(aggreads2)

#nb of detected presence per genus
sumpresgen<-rowSums(aggbin)
totpres<-sum(sumpresgen)

```

```{r possible graph, eval=FALSE}

######### Same by Class 

group_cols <- c("Class")  #We group by class. 
sample_cols <- names(merged)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- merged[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified class
aggclass <- result[complete.cases(result[, Class]), ]

newnames<- aggclass$Class
aggclass2 <- data.frame(aggclass[, -1], row.names = newnames)
colnames(aggclass2)<-colnames(aggclass[,2:180]) #to get rid of the newly added x before sample name

#Creation of the aggregated table of presence
aggbinclass<-aggclass2
aggbinclass[aggbinclass>0]<-1


#nb of reads per genus
sumgen<-rowSums(aggreads2)

#nb of detected presence per genus
sumpresgen<-rowSums(aggbin)
totpres<-sum(sumpresgen)

#then percent stacked barplot?

```

```{r asv nb}

####### Aggregation of table : get the nb of asv present for each genus

setDT(presence)

group_cols <- c("Phylum,Genus")  #We group by Genus. 
sample_cols <- names(presence)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- presence[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified Genus
aggasv <- result[complete.cases(result[, Genus]), ]

newnames<- aggasv$Genus
aggasv2 <- data.frame(aggasv[, -1:-2], row.names = newnames)
colnames(aggasv2)<-colnames(aggasv[,3:181]) #to get rid of the newly added x before sample name


```


## Multivariate analysis

### PCA analysis on the genus :

```{r PCA genera }

pcagen <- prcomp((aggreads2)) 

#pca on genera

# Eigenvalues
eig.val <- get_eigenvalue(pcagen)
eig.val[1:4,]
fviz_eig(pcagen)

#graph
fviz_pca_ind(pcagen,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.ind=list(contrib=10),
             repel=TRUE
             )

```

Blumeria genus is responsible for a majority of the variance on axis1 which represents 90.8% of the variance. Puccinia (wheat rust agent) is also well discribed on this axis, while all other genera are clustered together. Blumeria and Puccinia are opposed on axis 2 (7.7%) while all other genera cluster together around the origin.  

### PCA analysis on samples

```{r PCA samples}

pcasam <- prcomp(t(aggreads2))


# Eigenvalues
eig.val <- get_eigenvalue(pcasam)
eig.val[1:4,]
fviz_eig(pcasam)

fviz_pca_ind(pcasam,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )

fviz_pca_var(pcasam,
             col.var = "contrib", # Color by contributions to the principal axes
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var=list(contrib=(5)),
             repel = TRUE     # Avoid text overlapping
             )

#get the name of the top 10 genera that contribute most to the principal axes.
loading_scores <- pcasam$rotation[,1]
genera_scores <- abs(loading_scores) ## get the magnitudes
genera_score_ranked <- sort(genera_scores, decreasing=TRUE)
top10_genera <- names(genera_score_ranked[1:10])
 
pcasam$rotation[top10_genera,1:3] ## show the scores and sign



```

## Correspondance analysis on presence table

```{r ca on bintable}

#PCA
pcabin <- prcomp(t(aggbin))

# Eigenvalues
eig.val <- get_eigenvalue(pcabin)
eig.val[1:10,]
fviz_eig(pcabin)

fviz_pca_ind(pcabin,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )

fviz_pca_var(pcabin,
             col.var = "contrib", # Color by contributions to the principal axes
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var=list(contrib=20),
             repel = TRUE     # Avoid text overlapping
             )

# Contributions of variables to PC1
fviz_contrib(pcabin, choice = "var", axes = 1, top = 10)

```

Ce sont les genres "rares" qui sont mis en avant par l'ACP sur présence/absence. 

## Relation with the field

```{r patho}

#import field data on phytopathologies 
patho<-read.csv("notations_patho.csv",header=T)
colnames(patho)<-patho[2,]
tabpatho<-patho[-1:-2,]

#merge table
################# INSTEAD OF AGGREADS2 I CAN USE AGGASV2
taggreads<-t(aggreads2)
genotypes<-rownames(taggreads)
taggreads2<-cbind(genotypes,taggreads)
rownames(tabpatho)<-tabpatho[,1]
tabpatho2<-tabpatho[,-1]

readspatho<-merge(tabpatho2, taggreads, by=0)


nbgen<-rowSums(readspatho[,4:171])

pathonbgen<-cbind(nbgen,readspatho)

#pas de lien nb genre


```




## Diversity index

```{r diversity}



#Creation of frequency table necessary for shannon index

freqreads<-t(readspatho)
for (j in 1:171){
  tot<-sum(as.numeric(freqreads[4:1211,j]))
  for (i in 4:1211){
    val<-as.numeric(freqreads[i,j])
    freqreads[i,j]<-val/tot
  }
}

#Shannon diversity index
shindex<-c()
for (j in 1:171){
  sumline<-c()
  for (i in 4:1211){
    p<-as.numeric(freqreads[i,j])
    if (p>0) {
    sumline<-c(sumline, p*log(p))
    }
  }
  shindex<-c(shindex,-sum(sumline))
}
print(shindex)


#creation of table necessary for simpson index

simreads<-t(readspatho)
for (j in 1:171){
  tot<-sumsample[j]
  for (i in 4:1211){
    n<-as.numeric(simreads[i,j])
    if (n>0){
    simreads[i,j]<-n*(n-1)
    }
  }
}

#Simpson diversity index
simindex<-c()
for (j in 1:171){
  sumline<-c()
  for (i in 4:1211){
    ni<-as.numeric(simreads[i,j])
    if(ni>0){
    sumline<-c(sumline, ni)
    }
  }
  nn<-sumsample[j]
  simindex<-c(simindex,sum(sumline)/(nn*(nn-1)))
}
print(simindex)



```

Shannon index H' is an alpha-diversity index which account for the diversity of species (or as used here, genera) in a population (here, a sample). If a population is homogenous (with only one genus), H'=0. It is calculated as follows :
$H' = -\sum( p_i*ln(p_i))$

Simpson index $\lambda$ is another alpha-diversity index which account for the homogeneity of a population : the closer to 1, the more homogenous the population is. This index corresponds to the probability that two randomly selected individuals in a population belong to the same species. Compared to Shannon index, it gives more weight to abondant species, while Shannon index is more sensitive to the presence of rare species. It is calculated as follows:
$\lambda= \frac{\sum {n_i*(n_i-1)}}{N*(N-1)}$
