---
title: "Data analysis with removed 9 blumeria"
output: html_document
date: "2023-07-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data

```{r data, warning=FALSE, message=FALSE}
setwd("E:/Cours/SupAgro/Stage M1/Analyses/Romann/Data")

#packages
library(data.table)
library(dplyr)
library(ggplot2)
library(factoextra)
library(car)

rm(list=ls()) #to clear the memory if needed

##### DATA #####

#importation of reads table
seqtab_final<-readRDS("seqtab_final.rds")
tab_final<-t(seqtab_final)

tab_final2<-tab_final[,-4] #elimination of sample 104 because it looks like an error

errors<-c("AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGAATCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACTCTCAACGGGAATCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACTCTCAACGGGAATCGGGATGCGGCATCTGGTCACTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGATTCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGATTCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTAGACGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGACACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGATTCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGACCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGATTCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACGCAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGAATCGGAATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGCTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT",
          "AACGCAAGTTGCGCCCGAGGCCACTCGGCCGAGGGCACGCCTGCCTGGGCGTCACGCCAAAACACGCTCCCAACCACCCTCAACGGGATTCGGGATGCGGCATCTGGTCCCTCGTCTCTCAAGGGACGGTGGACCGAAGATTGGGTTGCCGGCGTACCGCGCCGGACACAGCGCATGGTGGGCGTCCTCGCTTTATCAATGCAGTGCATCCGGCGCGCAGCTGGCATTAT")
#these are the sequences attributed to Blumeria graminis that are actually wheat (according to a separate analysis and blasting)

seq<-rownames(tab_final2)#this vector contains all sequences
rowerrors<-c()#this one will contain the row nb of the sequences to discard
for (i in 1:length(seq)){
  for (a in 1:length(errors)){
    if (seq[i]==errors[a]){
      rowerrors<-c(rowerrors,i)
    }
  }
}

tab_final3<-tab_final2[-rowerrors,]
#We lost 9 asv and 2419 reads

colnames(tab_final3)<-paste0("EPO_", colnames(tab_final3))

#creation of binary (presence/absence) table
bin_final<-tab_final3
bin_final[bin_final>0]<-1

#importation of the taxa table
tax_final<-readRDS("tax_final.rds")


```

```{r merge}

merged<-merge(tax_final, tab_final3, by=0) #merged table with taxa and nb of reads

#binarytable
presence<-merge(tax_final, bin_final, by=0) #merged table with taxa and presence/absence

#sum of reads for an asv
readsum<-rowSums(merged[,9:187])
totreads<-sum(readsum) #8735570

#nb of samples with detection for an asv
presencesum<-rowSums(presence[,9:187])


```

9 sequences, that represented `2419` reads were deleted, because they were incorrectly attributed to Blumeria graminis while being wheat sequences.
The total number of reads among all asv is `8735570`. Genus Blumeria (powdery mildew agent) represents around `46%` of the reads and Puccinia (wheat rust) represents around 23% of the reads.

```{r hist}

#nb of reads per sample (histogram)
sumsample<-colSums(merged[,9:187])
hist(sumsample, col="steelblue", main="nb of reads per sample")
abline(v = mean(sumsample), col = 'red') #adding the mean vertical line

#nb of variants per sample (histogram)
presencesumsample<-colSums(presence[,9:187])
hist(presencesumsample, col="steelblue", main="nb of asv per sample")
abline(v = mean(presencesumsample), col = 'red') #adding the mean vertical line

```

```{r GROUP, message=FALSE}

#table with interesting values
values<-merged[,1:8]
values<-cbind(values,readsum,presencesum)

#to have a look at the data ! We can change the "group by" parameter
groupvalues <- values %>% group_by(Phylum, Genus) %>%
  summarise(total_reads = sum(readsum), maxnbpresence = max(presencesum), nbvar = n())

```

```{r aggregation for main tables}

####### Aggregation of reads table
setDT(merged)

group_cols <- c("Phylum,Genus")  #We group by Genus. 
sample_cols <- names(merged)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- merged[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified Genus
aggreads <- result[complete.cases(result[, Genus]), ]

newnames<- aggreads$Genus
aggreads2 <- data.frame(aggreads[, -1:-2], row.names = newnames)
colnames(aggreads2)<-colnames(aggreads[,3:181]) #to get rid of the newly added x before sample name

######Creation of the aggregated table of presence based on reads
aggbin<-aggreads2
aggbin[aggbin>0]<-1

####### Creation of the table nb of asv present for each genus

setDT(presence)

group_cols <- c("Phylum,Genus")  #We group by Genus. 
sample_cols <- names(presence)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- presence[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified Genus
aggasv <- result[complete.cases(result[, Genus]), ]

newnames<- aggasv$Genus
aggasv2 <- data.frame(aggasv[, -1:-2], row.names = newnames)
colnames(aggasv2)<-colnames(aggasv[,3:181]) #to get rid of the newly added x before sample name


```

```{r sums}

#nb of reads per genus
sumgen<-rowSums(aggreads2)

#nb of detected presence per genus
sumpresgen<-rowSums(aggbin)
totpres<-sum(sumpresgen)

data <- t(rbind(sumgen,sumpresgen))
data<-as.data.frame(data)
data<-top_n(data, 15, sumpresgen) 
knitr::kable(data,
caption = "Top 15 most present genera")

```


```{r var and reads}
tasv2<-as.data.frame(t(aggasv2))
treads2<-as.data.frame(t(aggreads2))
sumasv2<-rowSums(tasv2)
sumreads2<-rowSums(treads2)


plot(tasv2$g__Blumeria,treads2$g__Blumeria)
cor.test(tasv2$g__Blumeria,treads2$g__Blumeria)

plot(tasv2$g__Zymoseptoria,treads2$g__Zymoseptoria)
cor.test(tasv2$g__Zymoseptoria,treads2$g__Zymoseptoria)

plot(tasv2$g__Cladosporium,treads2$g__Cladosporium)
cor.test(tasv2$g__Cladosporium,treads2$g__Cladosporium)

plot(tasv2$g__Puccinia,treads2$g__Puccinia)
cor.test(tasv2$g__Puccinia,treads2$g__Puccinia)


```



```{r possible graph, eval=FALSE}

######### Same by Class 

group_cols <- c("Class")  #We group by class. 
sample_cols <- names(merged)[9:187]  # We want to apply the grouping to these columns (sample ones)

# We create new aggregated variable by sum
result <- merged[, lapply(.SD, sum), by = group_cols, .SDcols = sample_cols]

#We only keep identified class
aggclass <- result[complete.cases(result[, Class]), ]

newnames<- aggclass$Class
aggclass2 <- data.frame(aggclass[, -1], row.names = newnames)
colnames(aggclass2)<-colnames(aggclass[,2:180]) #to get rid of the newly added x before sample name

#Creation of the aggregated table of presence
aggbinclass<-aggclass2
aggbinclass[aggbinclass>0]<-1


#nb of reads per genus
sumgen<-rowSums(aggreads2)

#nb of detected presence per genus
sumpresgen<-rowSums(aggbin)
totpres<-sum(sumpresgen)

#then percent stacked barplot?

```

## Multivariate analysis

```{r PCA genus only, eval=FALSE}

pcagen <- prcomp((aggreads2)) 

#pca on genera

# Eigenvalues
eig.val <- get_eigenvalue(pcagen)
eig.val[1:4,]
fviz_eig(pcagen)

#graph
fviz_pca_ind(pcagen,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.ind=list(contrib=10),
             repel=TRUE
             )

#Blumeria genus is responsible for a majority of the variance on axis1 which represents 90.8% of the variance. Puccinia (wheat rust agent) is also well discribed on this axis, while all other genera are clustered together. Blumeria and Puccinia are opposed on axis 2 (7.7%) while all other genera cluster together around the origin.

```

### PCA analysis on nb of reads

```{r PCA on reads}

pcasam <- prcomp(t(aggreads2))


# Eigenvalues
eig.val <- get_eigenvalue(pcasam)
eig.val[1:4,]
fviz_eig(pcasam)

fviz_pca_ind(pcasam,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )

fviz_pca_var(pcasam,
             col.var = "contrib", # Color by contributions to the principal axes
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var=list(contrib=(5)),
             repel = TRUE     # Avoid text overlapping
             )

#get the name of the top 10 genera that contribute most to the principal axes.
loading_scores <- pcasam$rotation[,1]
genera_scores <- abs(loading_scores) ## get the magnitudes
genera_score_ranked <- sort(genera_scores, decreasing=TRUE)
top10_genera <- names(genera_score_ranked[1:10])
 
pcasam$rotation[top10_genera,1:3] ## show the scores and sign



```

Genera that represent the majority of reads (Blumeria, Puccinia) represent almost all the variance.


### PCA analysis on presence/absence

```{r PCA on presence}

#PCA
pcabin <- prcomp(t(aggbin))

# Eigenvalues
eig.val <- get_eigenvalue(pcabin)
eig.val[1:10,]
fviz_eig(pcabin)

fviz_pca_ind(pcabin,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )

fviz_pca_var(pcabin,
             col.var = "contrib", # Color by contributions to the principal axes
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var=list(contrib=20),
             repel = TRUE     # Avoid text overlapping
             )

# Contributions of variables to PC1
fviz_contrib(pcabin, choice = "var", axes = 1, top = 10)

```

"Rare" genera are the ones that are highlighted by the PCA on presence/absence.

## Relation with the field

```{r patho}

#import field data on phytopathologies 
patho<-read.csv("notations_patho.csv",header=T)
colnames(patho)<-patho[2,]
tabpatho<-patho[-1:-2,] #first lines are empty

#merge table
################# INSTEAD OF AGGASV2 I CAN USE AGGREADS2 to do it on nb of reads
tagg<-t(aggreads2)
rownames(tabpatho)<-tabpatho[,1]
tabpatho2<-tabpatho[,-1]

pathodata<-merge(tabpatho2, tagg, by=0)


nbgen<-rowSums(pathodata[,4:171])

pathonbgen<-cbind(nbgen,pathodata)

```

Results from the phytopathology observations performed in june are imported. They consist in an impact mark for powdery mildew and for wheat rust, for each genotype. We can have a look at this data to find a possible correlation.

```{r test patho}

#patho mark according to #nb of asv
plot(pathonbgen$g__Blumeria,pathonbgen$O誰dium, main="Powdery mildew mark according to nb of Blumeria variants")
cor.test(pathonbgen$g__Puccinia,as.numeric(pathonbgen$Rouille))
plot(pathonbgen$g__Puccinia,pathonbgen$Rouille, main="Wheat rust mark according to nb of Puccinia variants")

#patho mark according to #nb of genera (diversity)
plot(pathonbgen$nbgen,pathonbgen$O誰dium,main="Powdery mildew mark according to nb of fungi genera")

```

## Diversity index

```{r diversity}

#Creation of frequency table necessary for shannon index

freqreads<-t(pathodata)
for (j in 1:171){
  tot<-sum(as.numeric(freqreads[4:1211,j]))
  for (i in 4:1211){
    val<-as.numeric(freqreads[i,j])
    freqreads[i,j]<-as.numeric(val/tot)
  }
}

#Shannon diversity index
shindex<-c()
for (j in 1:171){
  sumline<-c()
  for (i in 4:1211){
    p<-as.numeric(freqreads[i,j])
    if (p>0) {
    sumline<-c(sumline, p*log(p))
    }
  }
  shindex<-c(shindex,-sum(sumline))
}

#creation of table necessary for simpson index

simreads<-t(pathodata)
for (j in 1:171){
  tot<-sumsample[j]
  for (i in 4:1211){
    n<-as.numeric(simreads[i,j])
    if (n>0){
    simreads[i,j]<-n*(n-1)
    }
  }
}

#Simpson diversity index
simindex<-c()
for (j in 1:171){
  sumline<-c()
  for (i in 4:1211){
    ni<-as.numeric(simreads[i,j])
    if(ni>0){
    sumline<-c(sumline, ni)
    }
  }
  nn<-sumsample[j]
  simindex<-c(simindex,sum(sumline)/(nn*(nn-1)))
}


```

Shannon index H' is an alpha-diversity index which account for the diversity of species (or as used here, genera) in a population (here, a sample). If a population is homogenous (with only one genus), H'=0. It is calculated as follows :
$H' = -\sum( p_i*ln(p_i))$  

Simpson index $\lambda$ is another alpha-diversity index which account for the homogeneity of a population : the closer to 1, the more homogenous the population is. This index corresponds to the probability that two randomly selected individuals in a population belong to the same species. Compared to Shannon index, it gives more weight to abondant species, while Shannon index is more sensitive to the presence of rare species. It is calculated as follows:
$\lambda= \frac{\sum {n_i*(n_i-1)}}{N*(N-1)}$  

These index take into account the diversity of taxa and the number of individuals within taxa. Here, we use the nb of variants instead of nb of individuals.

```{r test index}
hist(shindex, col="steelblue", main="Shannon index per population")
hist(simindex, col="steelblue", main="Simpson index per population")

plot(shindex,simindex)
cor.test(shindex,simindex)

plot(shindex,pathonbgen$O誰dium)
plot(simindex,pathonbgen$O誰dium)

```

